---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Read the data:
```{r results='hide'}
library(dplyr)
library(lattice)
data <- read.csv(file="activity.csv", header=TRUE)
```

Turn the interval column into factors
```{r}
data$interval <- factor(data$interval)
```

## What is mean total number of steps taken per day?

First, we sum the number of steps per day:
```{r}
per_day <- group_by(data, date)
stepsPerDay <- summarise(per_day, totalSteps=sum(steps))
```

Next, here is a histogram of the total number of steps taken each day, using 10 bins:

```{r}
hist(stepsPerDay$totalSteps, breaks=10)
```

The mean of the totalSteps per day is:
```{r}
mean(stepsPerDay$totalSteps, na.rm = TRUE)
```

The median of the totalSteps per day is:
```{r}
median(stepsPerDay$totalSteps, na.rm = TRUE)
```

## What is the average daily activity pattern?

We group the data by interval, and then find the mean (average) for each interval.
Then we create a plot of these data points.

```{r}
per_interval <- group_by(data, interval)
stepsPerInterval <- summarise(per_interval, averageSteps=mean(steps, na.rm=TRUE))
plot(stepsPerInterval$interval, stepsPerInterval$averageSteps ) 
lines(stepsPerInterval$interval, stepsPerInterval$averageSteps, type="l")
```

The 5-minute interval, that contains the maximum number of steps (average across all days in dataset) is:
```{r}
stepsPerInterval[which.max(stepsPerInterval$averageSteps),]
```

## Imputing missing values
1. The nr of missing values in the data set is:
````{r}
sum(is.na(data$steps))
```


2. We will fill in NA values, by using the mean of that day. If the mean of that day is also NA, we set it to be 0.

```{r}
#group data by date
per_day <- group_by(data, date)
#get average steps per day
averageStepsPerDay <- summarise(per_day, averageSteps=mean(steps))
#Set all NA values to be 0
averageStepsPerDay$averageSteps[is.na(averageStepsPerDay$averageSteps)] <- 0
filledData <- merge(data, averageStepsPerDay, by="date")
#use the averageSteps data to fill in any NA values
filledData$steps[is.na(filledData$steps)] <- filledData$averageSteps[is.na(filledData$steps)]
```

Now let's make a histogram again, for this new filled in data set.
```{r}
per_day2 <- group_by(filledData, date)
stepsPerDay2 <- summarise(per_day2, totalSteps=sum(steps))
hist(stepsPerDay2$totalSteps, breaks=10)
```

The mean of the totalSteps per day is now:
```{r}
mean(stepsPerDay2$totalSteps, na.rm = TRUE)
```

The median of the totalSteps per day is now:
```{r}
median(stepsPerDay2$totalSteps, na.rm = TRUE)
```


## Are there differences in activity patterns between weekdays and weekends?

```{r}
#add column with weekdays
filledData$day <- weekdays(as.Date(filledData$date,'%Y-%m-%d'))
#add column categorizing weekdays vs weekend
filledData$weekday <- ifelse(filledData$day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
#make this column a factor
filledData$weekday <- factor(filledData$weekday)
#group by interval and weekday
groups2 <- group_by(filledData, interval, weekday)
weekdayData <- summarise(groups2, averageSteps=mean(steps))


#create the plot
xyplot(averageSteps ~ interval | weekday, data=weekdayData,type='l', layout=c(1,2))


```

